---
title: "<Draft> Lab3: Reducing Crime"
subtitle: "w203 Lab3"
author: "Harith Elrufaie and Gaurav Desai"
output: pdf_document
---

# Exploratory Data Analysis

Perform an exploratory analysis to understand understand the determinants of crime and to generate policy suggestions that are applicable to local government.

## Setup

First, we load the necessary libraries.

```{r library}
library(stargazer)
library(corrplot)
```

Then we load the dataset, which is in the same directory as this RMD.

```{r data}
# Load the data 
rawCrimeData = read.csv("crime_v2.csv")
str(rawCrimeData)
```

The dataset contains 25 columns (variables) and 97 rows

## Data Quality/Clean-up

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
apply(!is.na(rawCrimeData[,]), MARGIN = 2, mean)
```

the dataset contains few NA that we'll need to fix before proceeding further. 

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
complete.cases(rawCrimeData)
crimeData <- rawCrimeData[complete.cases(rawCrimeData), ]
apply(!is.na(crimeData[,]), MARGIN = 2, mean)
```

now we're good to go.

------

## Univariate Variable Analysis of Key Variables

### crmrte: crimes committed per person

Looking at the histogram of crime per person, the distribution appear to be positelvey skewed. Applying `log()` on crime shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$crmrte, breaks=20, main = "Crimes Committed", cex.main=0.8)
hist(log(crimeData$crmrte), breaks=20, main = "Log Crimes Committed", cex.main=0.8)
```

### prbarr: probability of arrest

Looking at the histogram of arrest per person, the distribution appear to be positelvey skewed. Applying `log()` shows the histogram to appear *less* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$prbarr, breaks=20, main = "probability of Arrest", cex.main=0.8)
hist(log(crimeData$prbarr), breaks=20, main = "Log Probability of Arrest Log", cex.main=0.8)
```

### prbconv: probability of conviction

First, we'll need to concert the field from Factor to numeric for further analysis. 

```{r }
crimeData$prbconv <- as.numeric(as.character(crimeData$prbconv))
```

Looking at the histogram of probability of conviction, the distribution appear to be positelvey skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
crimeData$prbconv <- as.numeric(as.character(crimeData$prbconv))
hist(crimeData$prbconv, breaks=20, main = "Probability of Conviction",  cex.main=0.8)
hist(log(crimeData$prbconv), breaks=20, main = "Log Probability of Conviction",  cex.main=0.8)
```

### prbpris: of prison sentence

Looking at the histogram of probability of sentence, the distribution appear to be relatively normal. Applying `log()` shows the histogram to appear *less* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$prbpris, breaks=20, main = "Probability of Sentence",  cex.main=0.8)
hist(log(crimeData$prbpris), breaks=20, main = "Log Probability of Sentence",  cex.main=0.8)
```

### avgsen: avg. sentence, days

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear *more* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$avgsen, breaks=20, main = "Avg sentence in days", cex.main=0.8)
hist(log(crimeData$avgsen), breaks=20, main = "Log Avg sentence in days", cex.main=0.8)
```

### polpc: police per capita

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear *more* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$polpc, breaks=20, main = "Police per capita", cex.main=0.8)
hist(log(crimeData$polpc), breaks=20, main = "Log Police per capita", cex.main=0.8)
```

### density: people per sq. mile

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear *more* normally distributed, with one outlier.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$density, breaks=20, main = "People per sq. mile", cex.main=0.8)
hist(log(crimeData$density), breaks=20, main = "Log People per sq. mile", cex.main=0.8)
```

### taxpc: tax revenue per capita

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear slightly positevely skewed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$taxpc, breaks=20, main = "Tax revenue per capita", cex.main=0.8)
hist(log(crimeData$taxpc), breaks=20, main = "Log Tax revenue per capita", cex.main=0.8)
```

### taxpc: tax revenue per capita

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear slightly positevely skewed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
barplot(c(sum(crimeData$urban), sum(crimeData$west), sum(crimeData$central)),
  names.arg = c("Urban", "West", "Central"), main = "Part of the state counties are in", cex.main=0.8, cex.names = 0.7)
```

### pctmin80: perc. minority, 1980

Looking at the histogram of probability of sentence, the distribution appear to be slightly positevely skewed. Applying `log()` shows the histogram to appear negatively skewed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$pctmin80, breaks=20, main = "Perc. minority")
hist(log(crimeData$pctmin80), breaks=20, main = "Log Perc. minority")
```

### wcon: weekly wage, construction

Looking at the histogram of probability of sentence, the distribution appear to be slightly positevely skewed. Applying `log()` shows the histogram to appear normally distributed

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wcon, breaks=20, main = "Weekly wage (Construction)", cex.main=.8)
hist(log(crimeData$wcon), breaks=20, main = "Log Weekly wage (Construction)", cex.main=.8)
```

### wtuc: wkly wge, trns, util, commun

Looking at the histogram, the distribution appear to be normally distribbuted.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wtuc, breaks=20, main = "Weekly wage (trns/util/commun)", cex.main=.8)
```

### wtrd: wkly wge, whlesle, retail trade

Looking at the histogram, the distribution appear to be positvely skewed. Applying `log()` shows the histogram to appear normally distributed

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wtrd, breaks=20, main = "Weekly wage (whlesle, retail)", cex.main=.8)
hist(log(crimeData$wtrd), breaks=20, main = "Log Weekly wage (whlesle, retail)", cex.main=.8)
```

### wfir: wkly wge, fin, ins, real est

Looking at the histogram, the distribution appear to be normally distribbuted.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wfir, breaks=20, main = "Weekly wage (wge, fin, ins, real est)", cex.main=.8)
```

### wser: wkly wge, service industry

Looking at the histogram, the distribution appear to be positvely skewed. Applying `log()` shows the histogram to appear normally distributed with one outlier.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wser, breaks=20, main = "Weekly wage (service industry)", cex.main=.8)
hist(log(crimeData$wser), breaks=20, main = "Log Weekly wage (service industry)", cex.main=.8)
```

### wmfg: wkly wge, manufacturing

Looking at the histogram, the distribution appear to be slighly positvely skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wmfg, breaks=20, main = "Weekly wage (manufacturing)", cex.main=.8)
hist(log(crimeData$wmfg), breaks=20, main = "Log Weekly wage (manufacturing)", cex.main=.8)
```

### wfed: wkly wge, fed employees

Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wfed, breaks=20, main = "Weekly wage (fed employees)", cex.main=.8)
hist(log(crimeData$wfed), breaks=20, main = "Weekly wage (fed employees)", cex.main=.8)
```

### wsta: wkly wge, state employees

Looking at the histogram, the distribution appear to be slighly positvely skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wsta, breaks=20, main = "Weekly wage (state employees)", cex.main=.8)
hist(log(crimeData$wsta), breaks=20, main = "Log Weekly wage (state employees)", cex.main=.8)
```

### wloc: wkly wge, local gov emps

Looking at the histogram, the distribution appear to be slighly normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wloc, breaks=20, main = "Weekly wage (local gov employees)", cex.main=.8)
```

### mix: offense mix: face-to-face/other

Looking at the histogram, the distribution appear to be slighly positvely skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$mix, breaks=20, main = "Face-to-face/other", cex.main=.8)
hist(log(crimeData$mix), breaks=20, main = "Face-to-face/other", cex.main=.8)
```

### pctymle: percent young male

Looking at the histogram, the distribution appear to be positvely skewed. Applying `log()` shows the histogram to appear positvely skewed with one outlier.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$pctymle, breaks=20, main = "Percent Young Male", cex.main=.8)
hist(log(crimeData$pctymle), breaks=20, main = "Log Percent Young Male", cex.main=.8)
```


------

## Analysis of Key Relationships

It is very imperative to realize the relationship between crime rate and all the data available to us. We'll use `corrplot` to make the exploration of key relationships clearer.

```{r fig.align='center'}
corrplot(cor(crimeData[ , (names(crimeData) %in% c("crmrte", "prbarr", "prbconv", "prbpris", "avgsen", "polpc", "density", "taxpc", "west", "central", "uraban", "pctmin80", "wcon", "wtuc", "wtrd", "wfir", "wser", "wmfg", "wfed", "wsta", "wloc", "mix", "pctymle"))]), method="circle", type="upper")
```

The above plot indicates the following *positive* relationships with crime rate:
\begin{enumerate}
  \item Density (density).
  \item Tax revenue per capita (taxpc).
  \item All wage varibles.
  \item Young Male (pctymle)
\end{enumerate}

The above plot also indicates the following *negative* relationships with crime rate:
\begin{enumerate}
  \item Probabiliy of Arrest (prbarr)
  \item Probablity of Conviction (prbconv)
  \item West region of NC (west)
\end{enumerate}

### Crimes Committed per person (crmrte) & People per sq. (density)

As you can see from the correlation plot below, there is a positive linear relationship between crime rate and density. 

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(log(crimeData$density), log(crimeData$crmrte), 
    main="Crime Density vs Crime Rate", 
    xlab="Log Density",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$density)))
cor(crimeData$crmrte, crimeData$density)
```

### Crimes Committed per person (crmrte) & Tax revenue per capita (taxpc)

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(log(crimeData$taxpc), log(crimeData$crmrte), 
    main="Tax revenue per capita vs Crime Rate", 
    xlab="Tax revenue per capita",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$taxpc)))
cor(crimeData$crmrte, crimeData$taxpc)
```

### Crimes Committed per person (crmrte) & Wages

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(crimeData$wcon, log(crimeData$crmrte), 
    main="Wekly Wages (trns, util, commun) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ crimeData$wcon))

plot(log(crimeData$wtrd), log(crimeData$crmrte), 
    main="Wekly Wages (whlesle, retail trade) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wtrd)))

plot(crimeData$wfir, log(crimeData$crmrte), 
    main="Wekly Wages (fin, ins, real est) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ crimeData$wfir))

plot(log(crimeData$wser), log(crimeData$crmrte), 
    main="Wekly Wages (service industry) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wser)))

plot(log(crimeData$wmfg), log(crimeData$crmrte), 
    main="Wekly Wages (manufacturing) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wmfg)))

plot(log(crimeData$wfed), log(crimeData$crmrte), 
    main="Wekly Wages (fed employees) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wfed)))

plot(crimeData$wsta, log(crimeData$crmrte), 
    main="Wekly Wages (state employees) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ crimeData$wsta))

plot(log(crimeData$wloc), log(crimeData$crmrte), 
    main="Wekly Wages (local gov emps) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wloc)))
```

### Crimes Committed per person (crmrte) & Probabiliy of Arrest (prbarr)

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(log(crimeData$prbarr), log(crimeData$crmrte), 
    main="Probabiliy of Arrest vs Crime Rate", 
    xlab="Probabiliy of Arrest",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$prbarr)))
cor(crimeData$crmrte, crimeData$prbarr)
```

### Crimes Committed per person (crmrte) & Tax revenue per capita (prbconv)

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(log(crimeData$prbconv), log(crimeData$crmrte), 
    main="Probablity of Conviction vs Crime Rate", 
    xlab="Probablity of Conviction",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$prbconv)))
cor(crimeData$crmrte, crimeData$prbconv)
```

------

# Proposed Models

## Model 1: with only the explanatory variables

Using a combination of key positive and negative attributes to crime rate, we're recommending the following model:

$$crimeDeterm = \beta_0 + \beta_1 \cdot log(density) + \beta_2 \cdot log(taxpc) + \beta_3 \cdot log(prbarr) + \\
\beta_4 \cdot log(prbconv) + \beta_5 \cdot log(pctymle)$$

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
model1 <- lm(log(crmrte) ~ log(density) + log(taxpc) + log(prbarr) + log(prbconv) 
             + log(pctymle), data=crimeData)
summary(model1)
plot(model1)
```

## Model 2: with key explanatory variables and only covariates

In this model, we'll include the variables (avgsen, mix), as we think they will contribute to the accuracy of your results without introducing substantial bias.

$$crimeDeterm = \beta_0 + \beta_1 \cdot log(density) + \beta_2 \cdot log(taxpc) + \beta_3 \cdot log(prbarr) + \\
\beta_4 \cdot log(prbconv) + \beta_5 \cdot log(pctymle) + \beta_6 \cot log(avgsen) + \beta_7 \cot log(mix)$$

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
model2 <- lm(log(crmrte) ~ log(density) + log(taxpc) + log(prbarr) + log(prbconv) 
             + log(pctymle) + log(avgsen) + log(mix), data=crimeData)
summary(model2)
plot(model2)
```


## Model 3: includes the previous covariates, and most, if not all, other covariates

In this model, we'll include all the data available to us to demonstrate the robustness of of results to model specification. 

$$crimeDeterm = \beta_0 + \beta_1 \cdot log(density) + \beta_2 \cdot log(taxpc) + \beta_3 \cdot log(prbarr) + \\
\beta_4 \cdot log(prbconv) + \beta_5 \cdot log(pctymle) + \beta_6 \cdot log(avgsen) + \beta_7 \cdot log(mix) + \\
\beta_8 \cdot log(prbpris) + \beta_9 \cdot log(polpc) + \beta_10 \cdot log(pctmin80) + \beta_11 \cdot log(wcon)  \\
+ \beta_11 \cdot log(wtrd) + \beta_12 \cdot wfir + \beta_13 \cdot log(wser) + \beta_14 \cdot log(wmfg)  \\ 
+ \beta_15 \cdot log(wfed) + \beta_16 \cdot log(wsta) + \beta_17 \cdot wloc$$

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
model3 <- lm(log(crmrte) ~ log(density) + log(taxpc) + log(prbarr) + log(prbconv) 
             + log(pctymle) + log(avgsen) + log(mix) + log(prbpris) + log(polpc) 
             + log(pctmin80) + log(wcon) + log(wtrd) + wfir + log(wser) + log(wmfg)
             + log(wfed) + log(wsta) + wloc, data=crimeData)
summary(model3)
plot(model3)
```

## Models Regressions 

```{r } 
stargazer(model1, model2, model3, type = "text")
```