---
title: "<Draft> Lab3: Reducing Crime"
output: pdf_document
---

# Exploratory Data Analysis

Perform an exploratory analysis to understand how campaign characteristics are related to the candidate's debt. 

## Setup

First, we load the necessary libraries.

```{r library}
library(stargazer)
```

Then we load the dataset, which is in the same directory as this RMD.

```{r data}
# Load the data 
rawCrimeData = read.csv("crime_v2.csv")
str(rawCrimeData)
```

The dataset contains 25 columns (variables) and 97 rows

## Data Quality/Clean-up

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
apply(!is.na(rawCrimeData[,]), MARGIN = 2, mean)
```

the dataset contains few NA that we'll need to fix before proceeding further. 

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
complete.cases(rawCrimeData)
crimeData <- rawCrimeData[complete.cases(rawCrimeData), ]
apply(!is.na(crimeData[,]), MARGIN = 2, mean)
```

now we're good to go.

## Univariate Variable Analysis of Key Variables

### crmrte: crimes committed per person

Looking at the histogram of crime per person, the distribution appear to be positelvey skewed. Applying `log()` on crime shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$crmrte, breaks=20, main = "Crimes Committed", cex.main=0.8)
hist(log(crimeData$crmrte), breaks=20, main = "Log Crimes Committed", cex.main=0.8)
```

### prbarr: probability of arrest

Looking at the histogram of arrest per person, the distribution appear to be positelvey skewed. Applying `log()` shows the histogram to appear *less* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$prbarr, breaks=20, main = "probability of Arrest", cex.main=0.8)
hist(log(crimeData$prbarr), breaks=20, main = "Log Probability of Arrest Log", cex.main=0.8)
```

### prbconv: probability of conviction

First, we'll need to concert the field from Factor to numeric for further analysis. 

```{r }
crimeData$prbconv <- as.numeric(as.character(crimeData$prbconv))
```

Looking at the histogram of probability of conviction, the distribution appear to be positelvey skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
crimeData$prbconv <- as.numeric(as.character(crimeData$prbconv))
hist(crimeData$prbconv, breaks=20, main = "Probability of Conviction",  cex.main=0.8)
hist(log(crimeData$prbconv), breaks=20, main = "Log Probability of Conviction",  cex.main=0.8)
```

### prbpris: of prison sentence

Looking at the histogram of probability of sentence, the distribution appear to be relatively normal. Applying `log()` shows the histogram to appear *less* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$prbpris, breaks=20, main = "Probability of Sentence",  cex.main=0.8)
hist(log(crimeData$prbpris), breaks=20, main = "Log Probability of Sentence",  cex.main=0.8)
```

### avgsen: avg. sentence, days

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear *more* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$avgsen, breaks=20, main = "Avg sentence in days", cex.main=0.8)
hist(log(crimeData$avgsen), breaks=20, main = "Log Avg sentence in days", cex.main=0.8)
```

### polpc: police per capita

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear *more* normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$polpc, breaks=20, main = "Police per capita", cex.main=0.8)
hist(log(crimeData$polpc), breaks=20, main = "Log Police per capita", cex.main=0.8)
```

### density: people per sq. mile

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear *more* normally distributed, with one outlier.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$density, breaks=20, main = "People per sq. mile", cex.main=0.8)
hist(log(crimeData$density), breaks=20, main = "Log People per sq. mile", cex.main=0.8)
```

### taxpc: tax revenue per capita

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear slightly positevely skewed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$taxpc, breaks=20, main = "Tax revenue per capita", cex.main=0.8)
hist(log(crimeData$taxpc), breaks=20, main = "Log Tax revenue per capita", cex.main=0.8)
```

### taxpc: tax revenue per capita

Looking at the histogram of probability of sentence, the distribution appear to be positevely skewed. Applying `log()` shows the histogram to appear slightly positevely skewed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
barplot(c(sum(crimeData$urban), sum(crimeData$west), sum(crimeData$central)),
  names.arg = c("Urban", "West", "Central"), main = "Part of the state counties are in", cex.main=0.8, cex.names = 0.7)
```

### pctmin80: perc. minority, 1980

Looking at the histogram of probability of sentence, the distribution appear to be slightly positevely skewed. Applying `log()` shows the histogram to appear negatively skewed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$pctmin80, breaks=20, main = "Perc. minority")
hist(log(crimeData$pctmin80), breaks=20, main = "Log Perc. minority")
```

### wcon: weekly wage, construction

Looking at the histogram of probability of sentence, the distribution appear to be slightly positevely skewed. Applying `log()` shows the histogram to appear normally distributed

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wcon, breaks=20, main = "Weekly wage (Construction)", cex.main=.8)
hist(log(crimeData$wcon), breaks=20, main = "Log Weekly wage (Construction)", cex.main=.8)
```

### wtuc: wkly wge, trns, util, commun

Looking at the histogram, the distribution appear to be normally distribbuted.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wtuc, breaks=20, main = "Weekly wage (trns/util/commun)", cex.main=.8)
```

### wtrd: wkly wge, whlesle, retail trade

Looking at the histogram, the distribution appear to be positvely skewed. Applying `log()` shows the histogram to appear normally distributed

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wtrd, breaks=20, main = "Weekly wage (whlesle, retail)", cex.main=.8)
hist(log(crimeData$wtrd), breaks=20, main = "Log Weekly wage (whlesle, retail)", cex.main=.8)
```

### wfir: wkly wge, fin, ins, real est

Looking at the histogram, the distribution appear to be normally distribbuted.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wfir, breaks=20, main = "Weekly wage (wge, fin, ins, real est)", cex.main=.8)
```

### wser: wkly wge, service industry

Looking at the histogram, the distribution appear to be positvely skewed. Applying `log()` shows the histogram to appear normally distributed with one outlier.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wser, breaks=20, main = "Weekly wage (service industry)", cex.main=.8)
hist(log(crimeData$wser), breaks=20, main = "Log Weekly wage (service industry)", cex.main=.8)
```

### wmfg: wkly wge, manufacturing

Looking at the histogram, the distribution appear to be slighly positvely skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wmfg, breaks=20, main = "Weekly wage (manufacturing)", cex.main=.8)
hist(log(crimeData$wmfg), breaks=20, main = "Log Weekly wage (manufacturing)", cex.main=.8)
```

### wfed: wkly wge, fed employees

Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wfed, breaks=20, main = "Weekly wage (fed employees)", cex.main=.8)
hist(log(crimeData$wfed), breaks=20, main = "Weekly wage (fed employees)", cex.main=.8)
```

### wsta: wkly wge, state employees

Looking at the histogram, the distribution appear to be slighly positvely skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wsta, breaks=20, main = "Weekly wage (state employees)", cex.main=.8)
hist(log(crimeData$wsta), breaks=20, main = "Log Weekly wage (state employees)", cex.main=.8)
```

### wloc: wkly wge, local gov emps

Looking at the histogram, the distribution appear to be slighly normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$wloc, breaks=20, main = "Weekly wage (local gov employees)", cex.main=.8)
```

### mix: offense mix: face-to-face/other

Looking at the histogram, the distribution appear to be slighly positvely skewed. Applying `log()` shows the histogram to appear normally distributed.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$mix, breaks=20, main = "Face-to-face/other", cex.main=.8)
hist(log(crimeData$mix), breaks=20, main = "Face-to-face/other", cex.main=.8)
```

### pctymle: percent young male

Looking at the histogram, the distribution appear to be positvely skewed. Applying `log()` shows the histogram to appear positvely skewed with one outlier.

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
hist(crimeData$pctymle, breaks=20, main = "Percent Young Male", cex.main=.8)
hist(log(crimeData$pctymle), breaks=20, main = "Log Percent Young Male", cex.main=.8)
```

## Analysis of Key Relationships

### Crimes Committed per person (crmrte) & People per sq. (density)

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(log(crimeData$density), log(crimeData$crmrte), 
    main="Crime Density vs Crime Rate", 
    xlab="Log Density",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$density)))
cor(crimeData$crmrte, crimeData$density)
```

### Crimes Committed per person (crmrte) & Police per capita (polpc)

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(log(crimeData$polpc), log(crimeData$crmrte), 
    main="Crime Density vs Police per capita", 
    xlab="Poilice per capita",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$polpc)))
cor(crimeData$crmrte, crimeData$polpc)
```



### Crimes Committed per person (crmrte) & Percent of Young Male (pctymle)

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(log(crimeData$pctymle), log(crimeData$crmrte), 
    main="Crime Density vs Young Male", 
    xlab="Percent Young Male",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$pctymle)))
cor(crimeData$crmrte, crimeData$pctymle)
```

### Crimes Committed per person (crmrte) & Wages

```{r fig.align='center', fig.height=3, fig.width=3, fig.show='hold'}
plot(crimeData$wcon, log(crimeData$crmrte), 
    main="Wekly Wages (trns, util, commun) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ crimeData$wcon))

plot(log(crimeData$wtrd), log(crimeData$crmrte), 
    main="Wekly Wages (whlesle, retail trade) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wtrd)))

plot(crimeData$wfir, log(crimeData$crmrte), 
    main="Wekly Wages (fin, ins, real est) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ crimeData$wfir))

plot(log(crimeData$wser), log(crimeData$crmrte), 
    main="Wekly Wages (service industry) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wser)))

plot(log(crimeData$wmfg), log(crimeData$crmrte), 
    main="Wekly Wages (manufacturing) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wmfg)))

plot(log(crimeData$wfed), log(crimeData$crmrte), 
    main="Wekly Wages (fed employees) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wfed)))

plot(crimeData$wsta, log(crimeData$crmrte), 
    main="Wekly Wages (state employees) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ crimeData$wsta))

plot(log(crimeData$wloc), log(crimeData$crmrte), 
    main="Wekly Wages (local gov emps) vs Crime Rate", 
    xlab="Wekly Wages",
    ylab="Crimes Committed", cex.main=0.8)
abline(lm(log(crimeData$crmrte) ~ log(crimeData$wloc)))
```

# Proposed Models

## Model#1: with only the explanatory variables

$$crimeDeterm = \beta_0 + \beta_1 \cdot log(density) + \beta_2 \cdot log(prbarr) + \beta_3 \cdot log(prbconv) + \beta_4 \cdot log(mix) + \beta_5 \cdot log(pctymle)$$

```{r model1}
model1 <- lm(log(crmrte) ~ log(density) + log(prbarr) + log(prbconv) + log(mix) + log(pctymle), data=crimeData)
summary(model1)
plot(model1)
```

## Model#2: with key explanatory variables and only covariates

$$crimeDeterm = \beta_0 + \beta_1 \cdot log(density) + \beta_2 \cdot log(prbarr) + \beta_3 \cdot log(prbconv) + \beta_4 \cdot log(mix) + \beta_5 \cdot log(pctymle) + \beta_6 \cdot log(polpc)$$

```{r model2}
model2 <- lm(log(crmrte) ~ log(density) + log(prbarr) + log(prbconv) + log(mix) + log(pctymle) + log(polpc), data=crimeData)
summary(model2)
plot(model2)
```


## Model#3: includes the previous covariates, and most, if not all, other covariates

